---
title: "BMS5309_amoh0254_AT3"
author: "amoh0254"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BMS5309 Assessment Task 3

## Part A - Bulk RNA Sequencing Pipeline Codes:

### I. Bash Scripts

#### 1. Quality Control using FastQC and MultiQC:

```{bash, eval=FALSE}
#!/bin/bash

# Import modules
module load fastqc
module load multiqc

module avail fastqc
module avail multiqc

input_dir=~/analysis/data  
output_dir=~/analysis/fastqc_results  

# Create output directory if it doesn't already exist (-p allows 
# creation of nested  
# directories)  

mkdir -p "$output_dir"  
for file in "${input_dir}"/*.fastq  
do  
        echo "$file"  
        fastqc "$file" --outdir="$output_dir"  
done  
multiqc "$output_dir" -o "$output_dir"

```

#### 2. Adapter Trimming using Trimmomatic:

```{bash, eval=FALSE}
#!/bin/bash

# Import module
module load trimmomatic
module avail trimmomatic

input_dir=~/analysis/data
output_dir=~/analysis/trimmomatic
mkdir -p "$output_dir"
for file1 in "${input_dir}"/*_R1.fastq
do
file2="${file1/_R1.fastq/_R2.fastq}"

if [[ ! -f "$file2" ]]; then
echo "Warning: Pair file not found for $file1" continue
fi
# Extract sample base name (e.g., SRR975551)
name=$(basename "$file1" "_R1.fastq")
# Define output file names
out1="${output_dir}/${name}_1.trim_pe.fastq"
out2="${output_dir}/${name}_2.trim_pe.fastq"
out1se="${output_dir}/${name}_1.trim_se.fastq"
out2se="${output_dir}/${name}_2.trim_se.fastq"

echo "Running Trimmomatic for $name..."

java -jar "$TRIMMOMATIC_PATH"/trimmomatic-0.38.jar PE -threads 4 "$file1" "$file2" "$out1" "$out1se" "$out2" "$out2se" LEADING:3 TRAILING:3 SLIDINGWINDOW:4:15 MINLEN:50
done

```

#### 3. Post-trim Quality Control using FastQC and MultiQC:

```{bash, eval=FALSE}
#!/bin/bash

# Import modules
module load fastqc
module load multiqc

module avail fastqc
module avail multiqc

input_dir=~/analysis/trimmomatic 
output_dir=~/analysis/post_trim_fastqc_results 

# Create output directory if it doesn't already exist (-p allows 
# creation of nested  
# directories)  

mkdir -p "$output_dir"  
for file in "${input_dir}"/*.fastq  
do  
        echo "$file"  
        fastqc "$file" --outdir="$output_dir"  
done  
multiqc "$output_dir" -o "$output_dir"

```

#### 4. Read Alignment to Reference Genome using Hisat2:

```{bash, eval=FALSE}
#!/bin/bash  

# Import module
module load hisat2
module avail hisat2
 
reference_genome=~/bm26/reference_genomes/mouse/hisat2_ref/mouse_grcm39_index 
input_dir=~/analysis/trimmomatic
output_dir=~/analysis/trimmomatic_hisat2 
mkdir -p "$output_dir"  
for file1 in "${input_dir}"/*_1.trim_pe.fastq  
do   
file2="${file1/_1.trim_pe.fastq/_2.trim_pe.fastq}"   
if [[ ! -f "$file2" ]]; then   
echo "Paired file not found for: $file1"  
continue   
fi   
sample_name=$(basename "$file1"_1.trim_pe.fastq)  
output_sam="${output_dir}/${sample_name}.hisat2.sam"   
echo "Aligning sample: $sample_name"   
hisat2 -x "$reference_genome" -1 "$file1" -2 "$file2" -S "$output_sam"  
done 

```

#### 5. Mapping and Quantification of Transcripts using SAMtools and featureCounts:

```{bash, eval=FALSE}
#!/bin/bash  
# SAMtools

# Import module
module load samtools
module avail samtools

input_dir=~/analysis/trimmomatic_hisat2 
output_dir=~/analysis/aligned_sorted_bam  
mkdir -p "$output_dir"  
for file in "$input_dir"/*.sam  
do   
# Get the basename of the file minus the ".sam" ending   
name=$(basename "$file" .sam)   
echo "$name" 
# Convert SAM to BAM format   
samtools view -b "$file" > "$output_dir"/"$name".unsorted.bam   
# Sort BAM files   
samtools sort "$output_dir"/"$name".unsorted.bam > "$output_dir"/"$name".sorted.bam  
# Index BAM files   
samtools index "$output_dir"/"$name".sorted.bam   
# Generate summary stats   
samtools flagstat "$output_dir"/"$name".sorted.bam > "$output_dir"/"$name".flagstat   
# Delete the unsorted BAM
rm "$output_dir"/"$name".unsorted.bam 
done 

```

```{bash, eval=FALSE}
#!/bin/bash 
# featureCounts

# Import module
module load subread
module avail subread

reference_gtf="/home/amoh0254/bm26/reference_genomes/mouse/Mus_musculus.GRCm39.110.gtf"
bam_dir="/home/amoh0254/analysis/aligned_sorted_bam"  
output_dir="/home/amoh0254/analysis/counts/combined_counts.txt"
output_file="combined_counts.txt"  


mkdir -p "$output_dir"  

featureCounts -p -t exon -g gene_id -s 0 \
  -a "$reference_gtf" \
  -o "$output_dir/$output_file" \
  "$bam_dir"/*stimulated_R1.fastq_1.trim_pe.fastq_1.trim_pe.fastq.hisat2.sorted.bam
  
```

### II. R Scripts

#### 6. Library Size Normalization, 8. Differential Expression (DE):

-   Install/load required packages

```{r}
# Install required packages
if (!requireNamespace("BiocManager", quietly = TRUE)) { 
  install.packages("BiocManager") 
} 
if (!requireNamespace("edgeR", quietly = TRUE)) { 
  BiocManager::install("edgeR") 
} 
if (!requireNamespace("org.Mm.eg.db", quietly = TRUE)) { 
  BiocManager::install("org.Mm.eg.db") 
} 
if (!requireNamespace("GO.db", quietly = TRUE)) { 
  BiocManager::install("GO.db") 
} 
if (!requireNamespace("AnnotationDbi", quietly = TRUE)) { 
  BiocManager::install("AnnotationDbi") 
} 
if (!requireNamespace("tidyverse", quietly = TRUE)) { 
  install.packages("tidyverse") 
} 

# Load packages 
library(edgeR) 
library(org.Mm.eg.db) 
library(GO.db) 
library(AnnotationDbi) 
library(tidyverse) 

```

-   Read the count matrix

```{r}
counts <- read.table("~/bm26/week2_wshop1_data/sturner_time_course.tsv", header=T, 
                     sep="\t", stringsAsFactors = FALSE)

```

-   Inspect the data

```{r}
head(counts)

```

-   Prepare the count matrix

```{r}
#The important point is that the first column (X) contains the mouse gene symbols, while the 
#remaining 12 columns are your count data. Your first column is called X, so rename it to 
#GeneName. 

# Rename first column  
colnames(counts)[1] <- "GeneName" 

# Check the structure 
head(counts) 

dim(counts) 

# Check for duplicated gene names 
sum(duplicated(counts$GeneName)) 


# Make GeneName unique while keeping all rows 8
counts$GeneName <- make.unique(as.character(counts$GeneName), sep = "_")

# Set unique GeneName as row names 
rownames(counts) <- counts$GeneName

# Remove GeneName column 
counts$GeneName <- NULL

# Check 
head(counts) 

# Get the row names
head(rownames(counts)) 

#Check for duplicates 
anyDuplicated(rownames(counts))

```

-   Create sample information

```{r}
# Create DGEList 
dge <- DGEList(counts = counts) 

# Check the object 
dge

#need to assign each sample to its time point. 
group <- factor(c( 
  "3hr", "5hr", "24hr", "0hr", 
  "0hr", "3hr", "5hr", "24hr", 
  "3hr", "5hr", "24hr", "0hr" 
)) 
group 

#check that the groups match the sample names: 
data.frame(Sample = colnames(dge), Group = group ) 

dge$samples$group <- group 
dge$samples 

```

-   Filter lowly expressed genes

```{r}
#Filter low expressed genes 
# Calculate CPM 
cpm_values <- cpm(dge)

# Keep genes with CPM > 0.7 in at least 2 samples 
keep <- rowSums(cpm_values > 0.7) >= 2 

# Check how many genes will be retained 
table(keep)

dge <- dge[keep, , keep.lib.sizes = FALSE] 

# Check dimensions 
dim(dge) 

```

-   Normalise the data

```{r}
#perform TMM normalisation. 
dge <- calcNormFactors(dge) 

# Check normalisation factors 
dge$samples 

```

-   Define experimental design

```{r}
# create design matrix. There are four time points, treat them as categorical groups 
design <- model.matrix(~0 + group) 
colnames(design) <- levels(group) 
design 

```

-   Estimate dispersion

```{r}
#Estimate dispersion 
dge <- estimateDisp(dge, design) 

# View common dispersion 
dge$common.dispersion 

#look at the dispersion plot - helps visualise biological variability across genes 
plotBCV(dge)

```

-   Perform differential expression analysis

```{r}
#Fit the quasi-likelihood model 
fit <- glmQLFit(dge, design) 
fit 

#create contrast 
colnames(design) <- c("hr0", "hr3", "hr5", "hr24") 
design 

contrast_5hr_vs_3hr <- makeContrasts(hr5 - hr3,levels = colnames(design)) 
contrast_5hr_vs_3hr 

# logFC = expression at 5hr − expression at 3hr 
# Therefore: positive logFC → higher at 5hr, negative logFC → lower at 5hr  
# Perform the differential expression test 
qlf_5hr_vs_3hr <- glmQLFTest(fit, contrast = contrast_5hr_vs_3hr) 
qlf_5hr_vs_3hr

# extract all genes 
results_5hr_vs_3hr <- topTags(qlf_5hr_vs_3hr,n = Inf)$table 
head(results_5hr_vs_3hr)

```

-   Extract significant DE genes

```{r}
# Identify significant DE genes: FDR < 0.05, absolute log2FC ≥ 1 
sig_5hr_vs_3hr <- results_5hr_vs_3hr[results_5hr_vs_3hr$FDR < 0.05 & 
                                       abs(results_5hr_vs_3hr$logFC) >= 1,] 
head(sig_5hr_vs_3hr) 

#count them 
nrow(sig_5hr_vs_3hr) 

#Separate up- and down-regulated genes 
up_5hr_vs_3hr <- results_5hr_vs_3hr[results_5hr_vs_3hr$FDR < 0.05 & 
                                      results_5hr_vs_3hr$logFC >= 1,] 
down_5hr_vs_3hr <- results_5hr_vs_3hr[results_5hr_vs_3hr$FDR < 0.05 & 
                                        results_5hr_vs_3hr$logFC <= -1,] 
nrow(up_5hr_vs_3hr)  
nrow(down_5hr_vs_3hr)

#Sort the DE genes 
# most strongly up-regulated genes 
head(up_5hr_vs_3hr[order(-up_5hr_vs_3hr$logFC),])

# most strongly down-regulated 
head(down_5hr_vs_3hr[order(down_5hr_vs_3hr$logFC),])

#Save the results 
write.csv(results_5hr_vs_3hr,"DE_results_5hr_vs_3hr.csv",row.names = TRUE) 
write.csv(sig_5hr_vs_3hr,"Significant_DEG_5hr_vs_3hr.csv",row.names = TRUE) 

# Volcano plot 
# Add a significance classification 
results_5hr_vs_3hr$Significant[results_5hr_vs_3hr$FDR < 0.05 & 
                                 results_5hr_vs_3hr$logFC >= 1] <- "Up" 
results_5hr_vs_3hr$Significant[results_5hr_vs_3hr$FDR < 0.05 & 
                                 results_5hr_vs_3hr$logFC <= -1] <- "Down" 

#plot 
ggplot(results_5hr_vs_3hr, 
       aes( 
         x = logFC, 
         y = -log10(FDR), 
         colour = Significant 
       ) 
) + 
  geom_point(alpha = 0.6) + 
  geom_vline( 
    xintercept = c(-1, 1), 
    linetype = "dashed" 
  ) + 
  geom_hline( 
    yintercept = -log10(0.05), 
    linetype = "dashed" 
  ) + 
  theme_classic() + 
  labs( 
    title = "Differential Expression: 5hr vs 3hr", 
    x = "log2 Fold Change", 
    y = "-log10(FDR)" 
  ) 

```

-   Annotate genes

```{r}
# Go annotation with goanna: 
# Check the first few gene names 
head(rownames(dge)) 

# Create the annotation table 
test_annotation <- AnnotationDbi::select( 
  org.Mm.eg.db, 
  keys = rownames(sig_5hr_vs_3hr), 
  keytype = "SYMBOL", 
  columns = c("SYMBOL", "ENTREZID", "GENENAME") 
) 
head(test_annotation) 

#Check how many genes mapped 
sum(!is.na(test_annotation$ENTREZID))

length(unique(rownames(sig_5hr_vs_3hr)))

# This means 3470 of your 3768 significant genes have an Entrez ID mapping. 
#Create the Entrez ID list for goana() 
#This is important because your earlier error showed that your version of goana() expects a 
#list of character vectors. 
entrez_genes <- test_annotation$ENTREZID 
entrez_genes <- entrez_genes[!is.na(entrez_genes)] 
entrez_genes <- as.character(entrez_genes) 
head(entrez_genes)

length(entrez_genes) 

#Run goana() 
de_genes <- list(`5hr_vs_3hr` = entrez_genes) 
go_results <- goana(de_genes,species = "Mm") 
head(go_results) 

#Add FDR correction 
go_results$FDR <- p.adjust( go_results$P.5hr_vs_3hr, method = "BH" )

#sort by FDR 
go_results <- go_results %>% arrange(FDR) 
head(go_results) 

#Separate BP, MF and CC 
bp <- go_results %>% filter(Ont == "BP") 
mf <- go_results %>% filter(Ont == "MF") 
cc <- go_results %>% filter(Ont == "CC") 

#select significant terms 
sig_bp <- bp %>% filter(FDR < 0.05) 
sig_mf <- mf %>% filter(FDR < 0.05) 
sig_cc <- cc %>% filter(FDR < 0.05)

#number of significant GO terms 
nrow(sig_bp)

nrow(sig_mf) 

nrow(sig_cc) 

# top 10 Biological Processes 
head(sig_bp, 10) 

#save 
write.csv( 
  sig_bp, 
  "Biological_Process_5hr_vs_3hr_updated.csv", 
  row.names = TRUE 
) 
write.csv( 
  sig_mf, 
  "Molecular_Function_5hr_vs_3hr_upodated.csv", 
  row.names = TRUE 
) 
write.csv( 
  sig_cc, 
  "Cellular_Component_5hr_vs_3hr_updated.csv", 
  row.names = TRUE 
)

```

#### 7. Principal Component Analysis (PCA) based on normalized counts:

```{r}
#PCA check sample groups
install.packages("ggfortify")
library("ggfortify")
install.packages("ggrepel")
library("ggrepel")

logCPM <- cpm(dge,log=TRUE,prior.count=1)
sample_groups <- sub("X","",rownames(t(logCPM)))
sample_groups <- sub("_.*","",sample_groups)
dataPCA <- data.frame(
  sample = rownames(t(logCPM)),  
  grouping = sample_groups 
)
pca <- prcomp(t(logCPM))
autoplot(pca, data = dataPCA, colour = 'grouping', label = FALSE) +
  geom_text_repel(aes(label = dataPCA$sample), size = 3, max.overlaps = 20)

```

#### 9. Gene Set Enrichment of DE genes:

-   Gene Ontology (GO) Enrichment using GOanna

```{r}
#Gene Ontology (GO) Enrichment via goana 
go_results <- goana(de_genes, species = "Mm") 
go_results

go_results$FDR <- p.adjust(go_results$P.5hr_vs_3hr, method = "BH") 
go_results 

go_results <- go_results %>% arrange(FDR) 
go_results

bp <- go_results %>% filter(Ont == "BP") 
mf <- go_results %>% filter(Ont == "MF") 
cc <- go_results %>% filter(Ont == "CC") 
write.csv(bp %>% filter(FDR < 0.05), "Biological_Process_5hr_vs_3hr_camera.csv", 
          row.names = TRUE) 
write.csv(mf %>% filter(FDR < 0.05), "Molecular_Function_5hr_vs_3hr_camera.csv", 
          row.names = TRUE) 
write.csv(cc %>% filter(FDR < 0.05), "Cellular_Component_5hr_vs_3hr_camera.csv", 
          row.names = TRUE)

```

-   KEGG Pathway Enrichment using kegga

```{r}
#KEGG Pathway Enrichment via kegga
kegg_results <- kegga(de_genes, species = "Mm") 
kegg_results

kegg_results$FDR <- p.adjust(kegg_results$P.5hr_vs_3hr, method = "BH") 
sig_kegg <- kegg_results %>% filter(FDR < 0.05) %>% arrange(FDR) 
sig_kegg

write.csv(sig_kegg, "KEGG_Pathways_5hr_vs_3hr_camera.csv", row.names = TRUE) 

#Extract gene mapping from database 
all_genes_in_db <- keys(org.Mm.eg.db, keytype = "SYMBOL") 
go_mapping <- AnnotationDbi::select(org.Mm.eg.db, keys = all_genes_in_db, keytype = "SYMBOL", 
                     columns = c("GO", "ONTOLOGY")) 
go_mapping

#Filter strictly down to Biological Processes (BP)  
bp_mapping <- go_mapping %>% filter(ONTOLOGY == "BP" & !is.na(GO) & 
                                      !is.na(SYMBOL)) 
bp_mapping

#Create our structured gene list mapped by Gene Symbols to match DGEList matrix directly 
gene_sets_list <- split(bp_mapping$SYMBOL, bp_mapping$GO)

#Index matching positions relative to our active matrix rows 
indexed_sets <- ids2indices(gene_sets_list, identifiers = rownames(dge))

#Explanation for the above step : This step specifies which rows of the expression/count matrix belong to each gene set. Essentially, it maps the genes in each biological gene set to their corresponding row positions in the RNA-seq expression matrix. camera() needs to know which rows of your expression/count matrix belong to each gene set. 
# ids2indices() converts the gene names in each gene set into the row numbers corresponding to those genes in dge. 
#ids2indices() maps the genes in each biological gene set to their corresponding row positions in the RNA-seq expression matrix. This ensures that camera() tests the correct genes against the differential expression model. 
#Gene sets → ids2indices() → camera() → enriched pathways → FDR → biological interpretation

#Drop tiny categories to keep calculations statistically valid. Filter Out Very Small Gene Sets 
indexed_sets <- indexed_sets[sapply(indexed_sets, length) >= 5]

#Explanation of the above step: 
#• indexed_sets contains the gene sets after matching their genes to those present in the RNA-seq dataset.  
#• sapply(indexed_sets, length) calculates the number of genes from the RNA-seq dataset present in each gene set.  
#• >= 5 keeps only gene sets containing at least five genes from the dataset. #• Very small gene sets can produce unstable or unreliable enrichment statistics, so filtering them out improves the robustness and interpretability of the analysis. 

```

- CAMERA Enrichment

```{r}
#Run CAMERA across the target contrast design 
camera_results <- camera(dge, index = indexed_sets, design = design, contrast = 
                           contrast_5hr_vs_3hr) 
camera_results

sig_camera <- camera_results %>% filter(FDR < 0.05) 
sig_camera 

#Save result data matrix 
write.csv(sig_camera, "CAMERA_Competitive_Enrichment_5hr_vs_3hr.csv", row.names = 
            TRUE)

#Visualise the top pathways
#plots
#prepare the CAMERA results
# CAMERA DOT PLOT
camera_plot <- sig_camera %>%
  mutate(
    GO_ID = rownames(sig_camera),
    negLogFDR = -log10(FDR)) %>%
  arrange(FDR) %>%
  slice_head(n = 20)
# Convert GO IDs to readable GO terms
camera_plot$TERM <- Term(GOTERM[camera_plot$GO_ID])

# If a GO term cannot be found, keep the GO ID
camera_plot$TERM <- ifelse(
    is.na(camera_plot$TERM),
    camera_plot$GO_ID,
    camera_plot$TERM )

# Order pathways by significance
camera_plot$TERM <
  - factor(
    camera_plot$TERM,
    levels = rev(camera_plot$TERM) )
# Plot
ggplot(
  camera_plot,
  aes(
    x = negLogFDR,
    y = TERM,
    size = NGenes,
    shape = Direction
  )) +
  geom_point() +
  labs(
    title = "CAMERA Gene Set Enrichment",
    subtitle = "5 hr vs 3 hr",
    x = "-log10(FDR)",
    y = "GO Biological Process",
    size = "Number of genes",
    shape = "Direction"
  ) +
  
theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```

## Part B - Single-cell RNA Sequencing Pipeline Codes:

### I. Bash Scripts

#### 1. Cell Ranger:

-   Memory T-cells

```{bash, eval=FALSE}
#!/usr/bin/env bash
#SBATCH -p comp
#SBATCH --job-name="Cell_Ranger_memory"
#SBATCH --error=cellranger_err_memory.txt
#SBATCH --account=bm26
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=8
#SBATCH --mail-type=END,FAIL
#SBATCH --mem-per-cpu=64G
#SBATCH --time=0-02:00:00

module load cellranger/8.0.1
reference_path="/home/amoh0254/bm26_scratch/reference/refdata-gex-GRCm39-2024-A/"
#cd /home/amoh0254/bm26_scratch/amoh0254/scRNA/cellranger_results_WEEK6_02
cd /home/amoh0254/bm26_scratch/amoh0254/scRNA/cellranger_output_memory

fastq_path="/home/amoh0254/bm26_scratch/amoh0254/scRNA/WEEK6_scRNA_data_1M/memory/"
cellranger count --id="memory_M2" \
           --transcriptome="$reference_path" \
           --fastqs="$fastq_path" \
           --sample="memory_M2" \
           --create-bam=true 

#fastq_path="/home/amoh0254/bm26_scratch/amoh0254/scRNA/WEEK6_scRNA_data_1M/effector/"
#cellranger count --id="Effector_E2" \
#           --transcriptome=$reference_path \
#           --fastqs=$fastq_path \
#           --sample="Effector_E2" \
#          --create-bam=true 

#fastq_path="/home/amoh0254/bm26_scratch/amoh0254/scRNA/WEEK6_scRNA_data_1M/naive/"
#cellranger count --id="Naive_N2" \
#           --transcriptome=$reference_path \
#           --fastqs=$fastq_path \
#           --sample="Naive_N2" \
#           --create-bam=true 


echo "====== Job Resource Usage ======"
sacct -j $SLURM_JOB_ID --format=JobID,Elapsed,MaxRSS,MaxDiskRead,MaxDiskWrite
```

-   Effector T-cells

```{bash, eval=FALSE}
#!/usr/bin/env bash
#SBATCH -p comp
#SBATCH --job-name="Cell_Ranger_effector"
#SBATCH --error=cellranger_err_effector.txt
#SBATCH --account=bm26
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=8
#SBATCH --mail-type=END,FAIL
#SBATCH --mem-per-cpu=64G
#SBATCH --time=0-02:00:00

module load cellranger/8.0.1
reference_path="/home/amoh0254/bm26_scratch/reference/refdata-gex-GRCm39-2024-A/"
#cd /home/amoh0254/bm26_scratch/amoh0254/scRNA/cellranger_results_WEEK6_02
cd /home/amoh0254/bm26_scratch/amoh0254/scRNA/cellranger_output


#fastq_path="/home/amoh0254/bm26_scratch/amoh0254/scRNA/WEEK6_scRNA_data_1M/memory/"
#cellranger count --id="memory_M2" \
#           --transcriptome="$reference_path" \
#           --fastqs="$fastq_path" \
#           --sample="memory_M2" \
#           --create-bam=true 

fastq_path="/home/amoh0254/bm26_scratch/amoh0254/scRNA/WEEK6_scRNA_data_1M/effector/"
cellranger count --id="Effector_E2" \
           --transcriptome="$reference_path" \
           --fastqs="$fastq_path" \
           --sample="Effector_E2" \
           --create-bam=true 

#fastq_path="/home/amoh0254/bm26_scratch/amoh0254/scRNA/WEEK6_scRNA_data_1M/naive/"
#cellranger count --id="Naive_N2" \
#           --transcriptome=$reference_path \
#           --fastqs=$fastq_path \
#           --sample="Naive_N2" \
#           --create-bam=true 


echo "====== Job Resource Usage ======"
sacct -j $SLURM_JOB_ID --format=JobID,Elapsed,MaxRSS,MaxDiskRead,MaxDiskWrite
```

-   Naive T-cells

```{bash, eval=FALSE}
#!/usr/bin/env bash
#SBATCH -p comp
#SBATCH --job-name="Cell_Ranger_naive"
#SBATCH --error=cellranger_err_naive.txt
#SBATCH --account=bm26
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=8
#SBATCH --mail-type=END,FAIL
#SBATCH --mem-per-cpu=64G
#SBATCH --time=0-02:00:00

module load cellranger/8.0.1
reference_path="/home/amoh0254/bm26_scratch/reference/refdata-gex-GRCm39-2024-A/"
#cd /home/amoh0254/bm26_scratch/amoh0254/scRNA/cellranger_results_WEEK6_02
cd /home/amoh0254/bm26_scratch/amoh0254/scRNA/cellranger_output


#fastq_path="/home/amoh0254/bm26_scratch/amoh0254/scRNA/WEEK6_scRNA_data_1M/memory/"
#cellranger count --id="memory_M2" \
#           --transcriptome="$reference_path" \
#           --fastqs="$fastq_path" \
#           --sample="memory_M2" \
#           --create-bam=true 

#fastq_path="/home/amoh0254/bm26_scratch/amoh0254/scRNA/WEEK6_scRNA_data_1M/effector/"
#cellranger count --id="Effector_E2" \
#           --transcriptome=$reference_path \
#           --fastqs=$fastq_path \
#           --sample="Effector_E2" \
#          --create-bam=true 

fastq_path="/home/amoh0254/bm26_scratch/amoh0254/scRNA/WEEK6_scRNA_data_1M/naive/"
cellranger count --id="Naive_N2" \
           --transcriptome="$reference_path" \
           --fastqs="$fastq_path" \
           --sample="Naive_N2" \
           --create-bam=true 


echo "====== Job Resource Usage ======"
sacct -j $SLURM_JOB_ID --format=JobID,Elapsed,MaxRSS,MaxDiskRead,MaxDiskWrite
```

### II. R Scripts

#### 1. scRNA-seq Analysis Preliminary Steps using Seurat:

- Load the packages

```{r}
library(dplyr)
library(ggplot2)
library(Seurat)
library(patchwork)
library(harmony)
```

- Define paths

```{r}
# Change this to the directory containing your Cell Ranger results

base_dir <- "/home/amoh0254/bm26_scratch/amoh0254/scRNA/cellranger_output"

# Cell Ranger filtered matrix directories

effector_dir <- file.path(
  base_dir,
  "Effector_E2",
  "outs",
  "filtered_feature_bc_matrix"
)

memory_dir <- file.path(
  base_dir,
  "memory_M2",
  "outs",
  "filtered_feature_bc_matrix"
)

naive_dir <- file.path(
  base_dir,
  "Naive_N2",
  "outs",
  "filtered_feature_bc_matrix"
)

```

- Load Cell Ranger data

```{r}
# Effector

effector.data <- Read10X(data.dir = effector_dir)

# Memory

memory.data <- Read10X(data.dir = memory_dir)

# Naive

naive.data <- Read10X(data.dir = naive_dir)

```

- Create Seurat objects

```{r}
effector <- CreateSeuratObject(
  counts = effector.data,
  project = "Effector",
  min.cells = 3,
  min.features = 200
)

memory <- CreateSeuratObject(
  counts = memory.data,
  project = "Memory",
  min.cells = 3,
  min.features = 200
)

naive <- CreateSeuratObject(
  counts = naive.data,
  project = "Naive",
  min.cells = 3,
  min.features = 200
)

```

- Add sample information

```{r}
effector$sample <- "Effector"
memory$sample <- "Memory"
naive$sample <- "Naive"
```

- Calculate mitochondrial percentage

```{r}
# Mouse mitochondrial genes normally begin with "mt-"

effector[["percent.mt"]] <- PercentageFeatureSet(
  effector,
  pattern = "^mt-"
)

memory[["percent.mt"]] <- PercentageFeatureSet(
  memory,
  pattern = "^mt-"
)

naive[["percent.mt"]] <- PercentageFeatureSet(
  naive,
  pattern = "^mt-"
)

```

- Examine QC metrics

```{r}
VlnPlot(
  effector,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  ncol = 3
)

VlnPlot(
  memory,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  ncol = 3
)

VlnPlot(
  naive,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  ncol = 3
)

```

- QC scatter plots

```{r}
FeatureScatter(
  effector,
  feature1 = "nCount_RNA",
  feature2 = "nFeature_RNA"
)

FeatureScatter(
  memory,
  feature1 = "nCount_RNA",
  feature2 = "nFeature_RNA"
)

FeatureScatter(
  naive,
  feature1 = "nCount_RNA",
  feature2 = "nFeature_RNA"
)

```

- Filter low-quality cells

```{r}
effector <- subset(
  effector,
  subset = nFeature_RNA > 200 & percent.mt < 5
)

memory <- subset(
  memory,
  subset = nFeature_RNA > 200 & percent.mt < 5
)

naive <- subset(
  naive,
  subset = nFeature_RNA > 200 & percent.mt < 5
)

```

- Merge the three samples

```{r}
seurat_object <- merge(
  effector,
  y = list(memory, naive),
  add.cell.ids = c("Effector", "Memory", "Naive"),
  project = "Effector_Memory_Naive"
)

# Check number of cells from each sample

table(seurat_object$sample)
```

- Normalize the combined dataset

```{r}
seurat_object <- NormalizeData(
  seurat_object,
  normalization.method = "LogNormalize",
  scale.factor = 10000
)

```

- Find highly variable genes

```{r}
seurat_object <- FindVariableFeatures(
  seurat_object,
  selection.method = "vst",
  nfeatures = 2000
)

```

- Plot highly variable genes

```{r}
top_vf <- head(
  VariableFeatures(seurat_object),
  10
)

top_vf

plot1 <- VariableFeaturePlot(
  seurat_object
)

plot2 <- LabelPoints(
  plot = plot1,
  points = top_vf,
  repel = TRUE
)

plot2

```

- Scale the data

```{r}
all.genes <- rownames(seurat_object)

seurat_object <- ScaleData(
  seurat_object,
  features = all.genes
)

```

- PCA

```{r}
seurat_object <- RunPCA(
  seurat_object,
  features = VariableFeatures(
    object = seurat_object
  )
)

```

- Examine PCA

```{r}
print(
  seurat_object[["pca"]],
  dims = 1:5,
  nfeatures = 5
)


VizDimLoadings(
  seurat_object,
  dims = 1:2,
  reduction = "pca"
)

DimPlot(
  seurat_object,
  reduction = "pca",
  group.by = "sample"
)

```

#### 2. scRNA-seq Analysis using Seurat:

- UMAP

```{r}
seurat_object <- RunUMAP(
  seurat_object,
  dims = 1:10
)

```

- UMAP coloured by sample

```{r}
DimPlot(
  seurat_object,
  reduction = "umap",
  group.by = "sample"
)

```

- Find clusters

```{r}
seurat_object <- FindNeighbors(
  seurat_object,
  dims = 1:10
)

seurat_object <- FindClusters(
  seurat_object,
  resolution = 0.5
)

```

- UMAP coloured by cluster

```{r}
DimPlot(
  seurat_object,
  reduction = "umap",
  group.by = "seurat_clusters",
  label = TRUE
)

```

- UMAP showing both sample and cluster

```{r}
DimPlot(
  seurat_object,
  reduction = "umap",
  group.by = "sample"
)

DimPlot(
  seurat_object,
  reduction = "umap",
  group.by = "seurat_clusters",
  label = TRUE
)

```

- Save Seurat object

```{r}
saveRDS(
  seurat_object,
  file = "Effector_Memory_Naive_Seurat.rds"
)

```

#### 3. scRNA-seq Analysis using Harmony:

- Harmony integration

```{r}
seurat_object <- harmony:::RunHarmony.Seurat(
  object = seurat_object,
  group.by.vars = "sample",
  reduction.use = "pca",
  dims.use = 1:20
)

```

- Examine Harmony embedding

```{r}

DimPlot(
  seurat_object,
  reduction = "harmony",
  group.by = "sample"
)

```

- Find neighbours using Harmony

```{r}
seurat_object <- FindNeighbors(
  seurat_object,
  reduction = "harmony",
  dims = 1:20
)

```

- Find clusters

```{r}
seurat_object <- FindClusters(
  seurat_object,
  resolution = 0.5
)

```

- Run UMAP using Harmony

```{r}
seurat_object <- RunUMAP(
  seurat_object,
  reduction = "harmony",
  dims = 1:20
)

```

- UMAP coloured by sample

```{r}
DimPlot(
  seurat_object,
  reduction = "umap",
  group.by = "sample"
)

```

- UMAP coloured by cluster

```{r}
DimPlot(
  seurat_object,
  reduction = "umap",
  group.by = "seurat_clusters",
  label = TRUE
)

```

- UMAP split by sample

```{r}
DimPlot(
  seurat_object,
  reduction = "umap",
  group.by = "sample",
  split.by = "sample"
)

```

- Save integrated Seurat object

```{r}
saveRDS(
  seurat_object,
  file = "Effector_Memory_Naive_Harmony_Seurat.rds"
)

```
